Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer

### 2024-01-18 report/markdown/manuscript_figures_tables.Rmd
Figures, tables and methods for survival analysis

## Transcriptomic Collagen Score and prediction of biochemical relapse-free survival

Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of first- and second-order normalized, $\log_2$-transformed expression levels of the collagen-related genes of interest by Elastic Net multi-parameter Cox regression [@Simon2011; @Zou2005]. 
The Transcriptomic Collagen Score was computed as the linear predictor score of the Elastic Net Cox model with non-zero model coefficients presented in __Figure \@ref(fig:fig-coll-score)A__ and __Supplementary Table S15__.

Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training collective (TCGA) and validation data sets (remaining cohorts) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Supplementary Table 16__).

@@

```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Score and prediction of biochemical relapse-free survival.'}

@@

__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Score and prediction of biochemical relapse-free survival.__

_Transcriptional Collagen Score was established in the TCGA training cohort by multi-parameter Elastic Net Cox modeling of biochemical relapse-free survival as a function of first- and second-order normalized $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes._

_(A) Non-zero Elastic Net Cox model coefficients presented as hazard ratios (HR). Point color codes for marker status (blue: favorable, red: unfavorable), point size represents absolute value of HR._

>>>>

## Prediction of biochemical relapse-free survival by gradient boosted machine modeling

For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the Gradient Boosted Machine (GBM) algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Tables S15__). 
The Transcriptomic Collagen Score was computed as the linear predictor score of the GBM model. 
Importance of the explanatory variables for the predictive performance of the GBM model were measured with the relative influence method [@Friedman2001] and expressed as difference in the sum of squared errors ($\Delta SSE$, __Figure \@ref(fig:fig-coll-score)A__).

Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training pooled GEO collective and validation data sets (TCGA and DKFZ) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Figure \@ref(fig:fig-coll-score)B__, __Supplementary Table 16__). 
Of note, predictive performance of the GBM algorithm was excellent and reproducible in the investigated cohorts as compared with several other machine learning algorithms (e.g. Ridge Cox regression, Elastic Net or Support Vector Machines, __Supplementary Figure S9__). 
Differences in biochemical relapse-free survival in cancer patients stratified by tertiles of the Transcriptomic Collagen Scores were assessed by false discovery rate (FDR) corrected Peto-Peto test (__Figure \@ref(fig:fig-coll-score)C__).

@@

```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Model of biochemical relapse-free survival.'}

@@

__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Model of biochemical relapse-free survival.__

_Transcriptional Collagen Score was established in the pooled GEO training cohort by multi-parameter Gradient Boosted Machine (GBM) modeling of biochemical relapse-free survival as a function of normalized $log_2$-transformed and ComBat-adjusted expression levels of `r nrow(globals$genes_interest)` collagen-related genes._

_(A) Importance of explanatory variables measured by the gradient of sum of squared errors ($\Delta SSE$) during the training process attributed to particular collagen-related transcript levels._

### 2024-01-18 report/markdown/manuscript_supplement.Rmd
Figures, tables and methods for survival analysis

For multi-parameter Elastic Net Cox modeling, the packages _glmnet_ [@Friedman2010], _survival_ [@Therneau2000], _surviminer_ [@Kassambara2016] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used.

@@

## Survival analysis and transcriptional collagen score

Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of normalized $log_2$-transformed expression of the collagen-related genes of interest by multi-parameter Elastic Net Cox modeling [@Simon2011; @Zou2005]. 
Of note, both first and second order terms were included in the Elastic Net model to account for possible non-linear association of gene expression with survival probability. 
The optimal values of the shrinkage parameter $\lambda$ was found by tuning in 200 repeats 10-fold cross-validation and corresponded to the model with the lowest deviance (function `cv.glmnet()`, package _glmnet_) [@Friedman2010]. 
The Elastic Net model was trained for the TCGA data set for the tuned $\lambda$ parameter with the `glmnet()` function from the _glmnet_ package. 
The Transcriptional Collagen Score was defined as the linear predictor score of the Elastic Net model and was computed for the training TCGA cohort and the validation collectives GSE54460, GSE70768, GSE70769, GSE220095 and DKFZ with the `predict()` method. 
Gene expression values used for calculation of the Transcriptional Collagen Score, i.e. non-zero coefficients of the Elastic Net model, are presented as hazard ratios (HR) in __Figure 3A__ and __Supplementary Table S\@ref(tab:tab-coll-score)__.

Performance of the Transcriptional Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-coll-survival)__). 
Statistical significance of differences in biochemical relapse-free survival between patients stratified by tertiles of the Transcriptional Collagen Score was determined by Peto-Peto test (functions `surv_fit()` and `surv_pvalue()`, package _survminer_, __Figure 3C__).

@@

```{r tab-coll-score, tab.cap = 'Transcriptomic collagen score was established by Elastic Net Cox modeling of biochemical relapse-free survival in the TCGA cohort. Hazard ratios (HR) of the member genes are presented.'}

flextable::flextable(suppl_paper_tbl$coll_score) %>% 
  set_widths(c(3, 3)) %>% 
  italic(1) %>% 
  theme_vanilla

@@

```{r tab-coll-survival, tab.cap = 'Performance of the transcriptomic collagen score at prediction of biochemical relapse-free survival in the TCGA training cohort and validation collectives.'}

flextable::flextable(suppl_paper_tbl$coll_survival) %>% 
  set_widths(c(2.8, 3, 2.7, 2.7, 3, 3)) %>% 
  merge_v(1) %>% 
  theme_vanilla

@@

_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts. P values were corrected for multiple testing with the false discovery rate (FDR) method. Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)_

@@

>>>>

Multi-parameter modeling of biochemical relapse-free survival was done with the packages 
_glmnet_ [@Friedman2010] (Ridge, Elastic Net and LASSO cox regression), 
_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
_survival_ [@Therneau2000], 
_surviminer_ [@Kassambara2016] and 
[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used. 
Univariable analysis of relapse-free survival was accomplished with [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer').

@@

## Multi-parameter survival modeling and transcriptional collagen score

For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__). 
The tuning was performed with an in-house developed script. 
The GBM model was trained in the pooled GEO cohort with the optimal tuning parameter set with the `gbm()` function (package _gbm_). 
The Transcriptomic Collagen Score was defined as the linear predictor score of the GBM model and was computed for the best model iteration (function `gbm.perf(method = 'cv')`) with the `predict()` method and expression data from the training pooled GEO cohort and the TCGA and DKFZ test collectives. 
Importance of the explanatory variables for the predictive performance of the GBM model in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables ($\Delta SSE$, __Figure 3A__).

Performance of the Transcriptional Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-model-survival)__). 
Of note, as compared with several other machine learning algorithms (Ridge, Elastic Net, and LASSO Cox regression, SVM and Random Forest) [@Friedman2010; @Fouodo2018; @Ishwaran2008; @Ishwaran2022a], the GBM model demonstrated an excellent performance and the highest reproducibility at prediction of biochemical relapse-free survival in the investigated cohorts (__Supplementary Figure S\@ref(fig:fig-surv-algos)__, __Supplementary Table S\@ref(tab:tab-model-survival)__). 
Statistical significance of differences in biochemical relapse-free survival between patients stratified by tertiles of the Transcriptional Collagen Score was determined by Peto-Peto test (functions `surv_fit()` and `surv_pvalue()`, package _survminer_, __Figure 3C__).

## Survival analysis for single collagen-related transcripts

The univariable analysis of biochemical relapse-free survival for single collagen-related genes was done in the pooled GEO, TCGA and DKFZ data sets described above for the multi-parameter survival modeling. 
For each of the collagen-related genes, cancer patients were classified as high and low expressors by an expression cutoff corresponding to the largest difference in biochemical relapse-free survival measured by FDR-corrected Mentel-Henszel test (function `find_cutoff()`, package [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer), minimal expression strata size set to 25% of the observations). 
Hazard ratio (HR) with 95% confidence intervals for relapse risk in high expressors as compared with low expressors was calculated with the canonical _survival_ package functions _coxph()_, _coef_ and _confint()_. 
Genes associated with unfavorable prognosis were defined by pFDR < 0.05 and HR > 1 for the high vs low expressor comparison. 
Genes associated with favorable prognosis were defined by pFDR < 0.05 and HR < 1 (__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano)__). 
Common survival markers were defined as genes significantly associated with survival in all cohorts (__Supplementary Figure S\@ref(fig:fig-surv-uni-forest)__ - __S\@ref(fig:fig-surv-uni-detal3)__, __Supplementary Table S\@ref(tab:tab-rfs-univariable)__).

@@

```{r tab-tuning, tab.cap = 'Selection of the optimal parameters of machine learining survival models of biochemical relapse-free survival with expression of the collagen-related genes. The selection process was accomplished by cross-validation tuning in the pooled GEO cohort.'}

flextable::flextable(suppl_paper_tbl$tuning) %>% 
  set_widths(c(3.5, 4, 3, 3)) %>% 
  merge_v(1:2, combine = TRUE) %>% 
  footnote(1, 1, 
           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'),
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CV: cross-validation.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-model-survival, tab.cap = 'Performance of machine learning models at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}

flextable::flextable(suppl_paper_tbl$model_survival) %>% 
  set_widths(c(3.5, 2.7, 3, 2.8, 2.8)) %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('pooled GEO: GSE54460, GSE70768, GSE70769, and GSE220095'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```{r tab-rfs-univariable, tab.cap = 'Results of univariable analysis of biochemical relapse-free survival with expression of the collagen-related genes. Prostate cancer patients were stratified by expression cutoffs corresponding to the largest difference in survival assessed by Mentel-Henszel test. Genes found to be significantly associated with the survival in all analyzed cohorts (pooled GEO, TCGA and DKFZ) are presented. The table is available in a supplementary Excel file.'}

suppl_paper_tbl$rfs_univariable[1, 1] %>% 
  mutate( `Association with survival` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts. P values were corrected for multiple testing with the false discovery rate (FDR) method. Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)._

@@

```{r fig-surv-algos, fig.width = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$h, fig.cap = 'Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}

suppl_paper_fig$surv_algos$plot

__Supplementary Figure S\@ref(fig:fig-surv-algos). Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.__

_Biochemical relapse-free survival was modeled with normalized, $log_2$-transformed and ComBat-adjusted expression levels of the collagen-related genes by various machine learning algorithms (Ridge Cox regression, Elastic Net Cox regression, LASSO Cox regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]). Their performance at prediction of biochemical relapse-free survival in the training pooled GEO cohort, and the test TCGA and DKFZ collectives was assessed by Harrell's concordance index (C-index) and integrated Brier scores (IBS). High C-index values indicate good concordance between the observed and predicted survival. Low values of IBS indicate good model calibration._
_C-index and IBS values are presented in scatter plots. Each point represents a single algorithm. Values of C-index and IBS expected for a model predicting survival at random are visualized as dashed lines._

```{r fig-surv-uni-volcano, fig.width = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.'}

suppl_paper_fig$surv_uni_volcano$plot

__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.__

_Cancer patients were stratified by cutoffs of normalized expression for each collagen-related genes which corresponded to the largest differences in biochemical relapse-free survival assessed by false discovery rate (FDR) corrected Mentel-Henszel test into high and low expressors for particular genes. Relative risk of biochemical relapse in high expressors as compared with low expressors was modeled by Cox proportional hazard regression and expressed as hazard ratio (HR). Unfavorable survival markers were defined by pFDR < 0.05 and HR > 1. Favorable survival markers were defined by pFDR < 0.05 and HR < 1. HR and pFDR values are presented in Volcano plots. Each point represents a single collagen-related gene. Significance and sign of association with the relapse risk are color coded. Significant genes are labeled with their symbols. Dashed lines indicate the significance and regulation cutoffs._

```{r fig-surv-uni-forest, fig.width = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$h, fig.cap = 'Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.'}

suppl_paper_fig$surv_uni_forest$plot

__Supplementary Figure S\@ref(fig:fig-surv-uni-forest). Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Favorable survival markers were defined by pFDR < 0.05 and HR < 1. HR with 95% confidence intervals (CI) for unfavorable and favorable markers shared by all investigated cohorts are presented in a Forest plot._

```{r fig-surv-uni-detail1, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.'}

suppl_paper_fig$surv_uni_detail1$plot

__Supplementary Figure S\@ref(fig:fig-surv-uni-detail1). Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL1A1, COL3A1, and COL5A2 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._

```{r fig-surv-uni-detail2, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.'}

suppl_paper_fig$surv_uni_detail2$plot

__Supplementary Figure S\@ref(fig:fig-surv-uni-detail2). Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL11A1, COL11A2, LOXL1, and SERPINH1 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._

```{r fig-surv-uni-detal3, fig.width = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.'}

suppl_paper_fig$surv_uni_detal3$plot

__Supplementary Figure S\@ref(fig:fig-surv-uni-detal3). Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.__

_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1. Differences in biochemical relapse-free survival for the COL4A6, DST, LAMB3, and PCOLCE2 genes identified as favorable markers in all investigated cohorts are shown in Kaplan-Meier plots. Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions. HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._
